---
title: "Exercise - GLM"
author: "Dr. F.J. Rodenburg"
format: html
editor: visual
---

## Description

In this fictional study, $n=50$ plants were measured in terms of stem size (mm). Then, the number of herbivore insects on each plant were counted. Investigate whether there are more insects on longer plants.

## Dependencies

If you use any packages, include them here, not where they are (first) used. This avoids multiple calls to the same package. You can add a short comment after each package explaining what it's used for.

```{r}

```

## Reading the data

Read the data and perform at least some data quality checks:

-   Do the dimensions (rows, columns) match what you expect? A common problem is when the delimiter is not recognized, in this case the number of columns.

-   Are numbers recognized as numeric?

-   Are categorical variables stored as `factor`, or `character`? In case of the latter, convert them to `factor`.

-   Are there any missing values? Any unrealistic values?

```{r}

```

If you note anything interesting, write this down over here, not in the code chunk.

## Visualizing the data

A simple visualization often goes a long way towards understanding the data. Create a figure with:

-   Stem size on the $x$-axis and the number of insects on the $y$-axis.

-   Informative axes.

```{r}

```

If the visualization reveals anything important, write that down as well:

## Fitting the model

This part does not produce any output. However, any warnings or errors that occur may reveal issues with model specification.

```{r}

```

## Model diagnostics

Always perform visual diagnostics on models used for inference. In case of GLMs, pay particular attention to overdispersion. And see what relevant information you can get with:

-   The `DHARMa` package;

-   The regression table (next section).

```{r}

```

## Interpreting the model output

```{r}

```

Write down what you see in the regression table, and what to make of it.

## Visualize the estimated relationship

This is a nice addition for any report/presentation. Try using the `plot_model` function from `sjPlot` to visualize the fit.

```{r}

```
